---
title: "Get Started"
output: rmarkdown::html_vignette
author: "Rafael Lopes & Leonardo Bastos"
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = F, 
  message = F, 
  echo = T
)
```

## First example on LazyData

When the package is loaded it provides a Lazy Data file, `sariBH`, it is a anonymized records of Severe Acute Respiratory Illness notified in the city of Belo Horizonte, since March 2020 to April 2022. To load it basically write:

```{r data-bh}
library(nowcaster)
# Loading Belo Horizonte SARI data set
data(sragBH)
```

And we take a look on the data:

```{r lookup}
head(sragBH)
```

It is a data.frame with 7 variables and 65,404 observations. We will make use of only the first two columns, "DT_SIN_PRI" (date of onset symptoms) and "DT_DIGITA" (recording date) as well the column "Idade" (age in years) to make age structured nowcasting.

The call of the function is straightforward, it simply needs a data set as input, here the `LazyData` loaded in the namespace of the package. The function has 3 mandatory parameters, `dataset` for the parsing of the data set to be nowcasted, `date_onset` for parsing the column name which is the date of onset of symptoms and `date_report` which parses the column name for the date of report of the cases. Here this columns are "DT_SIN_PRI" and "DT_DIGITA", respectively. 

```{r no_age}
nowcasting_bh_no_age <- nowcasting_inla(dataset = sragBH, 
                                        date_onset = "DT_SIN_PRI", 
                                        date_report = "DT_DIGITA", 
                                        data.by.week = T)
head(nowcasting_bh_no_age$total)
```

This calling will return for the first element the nowcasting estimate and its Confidence Interval (CI) for two different Credible interval, `LIb` and `LSb` are the max and min CI, respectively, with credibility of 50% and `LI` and `LS` are the max and min CI, respectively, with credibility of 95%.

On the second element it returns the data to be grouped and summarized to give the epidemic curve, we can take a look on this element. 

```{r epi-curve-plot}
library(ggplot2)
library(dplyr)

dados_by_week <- nowcasting_bh_no_age$data |> 
  dplyr::group_by(dt_event) |> 
  dplyr::reframe(
    observed = sum(Y, na.rm = T)
  ) |>
  dplyr::filter(dt_event >= max(dt_event)-270)

dados_by_week |> 
  ggplot()+
  geom_line(data = dados_by_week, 
            aes(dt_event, 
                y = observed, 
                col = 'Observed'))+
  theme_bw()+
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) +
  scale_color_manual(values = c('grey50', 'black'), 
                     name = '')+
  scale_x_date(date_breaks = '2 weeks', 
               date_labels = '%V/%y', 
               name = 'Date in Weeks')+
  labs(x = '', 
       y = 'Nº Cases')
```


After this element is groped by and summarized by the onset of symptoms date, here `DT_SIN_PRI`, it is the epidemiological curve observed. To example how the estimate compares with the observed curve, we plot the estimate and the epidemiological curve all together.

```{r no_age_plot}

nowcasting_bh_no_age$total |> 
  ggplot(aes(x = dt_event, y = Median, 
             col = 'Nowcasting')) +
  geom_line(data = dados_by_week, 
            aes(y = observed, 
                col = 'Observed'))+
  geom_ribbon(aes(ymin = LI, ymax = LS, col = NA), 
              alpha = 0.2, 
              show.legend = F)+
  geom_line()+
  theme_bw()+
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) +
  scale_color_manual(values = c('grey50', 'black'), 
                     name = '')+
  scale_x_date(date_breaks = '2 weeks', 
               date_labels = '%V/%y', 
               name = 'Date in Weeks')+
  labs(x = '', 
       y = 'Nº Cases')
```

This is an example were the estimate was done without considering any type of structure in data, which is the first assumption for the nowcasting.
