---
title: "Nowcasting on Structured data"
output: rmarkdown::html_vignette
author: "Rafael Lopes & Leonardo Bastos"
vignette: >
  %\VignetteIndexEntry{Nowcasting on Structured data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = F,
  message = F,
  echo = T
)
```

## TL;DR

This is a more length article explaining the workings and the choices behind the nowcasting model and its advantages and limitations.

## As Before...

As in the [Get Started](https://covid19br.github.io/nowcaster/articles/nowcaster.html) we start by loading the package and its lazy data, by:

```{r data-bh}
library(nowcaster)
# Loading Belo Horizonte SARI data set
data(sragBH)
```

## Non-structured data

The [Get Started](https://covid19br.github.io/nowcaster/articles/nowcaster.html) example is done on non-structured data, here we give a more detailed description of this type of data and how it can change the nowcasting.

When we call the `nowcasting_inla()` function, it has by default the parameterization to take the data and estimate with a non-structured data form. The estimate fits a negative binomial distribution, $NegBinom(\lambda_{t,d}, \phi)$, to the cases count at time $t$ with delay $d$, $\phi$ is the dispersion parameter. The rate $\lambda_{t,d}$ is then parametric in a log-linear format by a constant term added by structured delay random effects and structured time random effects.
This is the whole premise of nowcasting, we model the case counts at time $t$ given the case count at that time and past values as well as given the delay distribution of that count.

Hence, the model is given by the following:

$$\begin{equation}
Y_{t,d} \sim NegBinom(\lambda_{t,d}, \phi) \\
\log(\lambda_{t,d}) = \alpha + \beta_t + \gamma_d \\
\beta_t := u_t - u_{(t-1)} - u_{(t-2)} \sim N(0,\tau_{\beta}); \gamma_d := u_d - u_{(d-1)} \sim N(0, \tau_{\gamma})\\
t=1,2,\ldots,T \\ 
d=1,2,\ldots,D \\
\end{equation}$$

Where the intercept $\alpha$ follows a Gaussian distribution with a very large variance, $\beta_t$ follows a second order random walk with precision $\tau_\beta$, while $\gamma_d$ a first-order random walk with precision $\tau_\gamma$. The model is then completed by INLA default prior distributions for $\phi$, $\tau_\beta$, and $\tau_\gamma$. See [`nbinomial`](https://inla.r-inla-download.org/r-inla.org/doc/likelihood/likelihood-example.pdf), [`rw1`](https://inla.r-inla-download.org/r-inla.org/doc/latent/rw1.pdf) and [`rw2`](https://inla.r-inla-download.org/r-inla.org/doc/latent/rw1.pdf) INLA help pages.

The call of the function is straightforward, it simply needs a data set as input, here the `LazyData` loaded in the namespace of the package. The function has 3 mandatory parameters, `dataset` to parse the data set to be nowcasted, `date_onset` for parsing the column name which is the date of onset of symptoms and `date_report` which parses the column name for the date of report of the cases. Here this columns are "DT_SIN_PRI" and "DT_DIGITA", respectively. Need for two dates is due to we are modeling the delay as part of the nowcasting estimation, usually nowcasting models assuming a delay distribution and apply it to the cases observed.

```{r no_age}
nowcasting_bh_no_age <- nowcasting_inla(dataset = sragBH, 
                                        date_onset = DT_SIN_PRI, 
                                        date_report = DT_DIGITA)
head(nowcasting_bh_no_age$total)
```

The above calling will return only the nowcasting estimate and its Confidence Interval (CI) for two different credibility levels, `LIb` and `LSb` are the max and min CI, respectively, with credibility of 50% and `LI` and `LS` are the max and min CI, respectively, with credibility of 95%.

The `nowcasting_inla` has the option to return the curve on which the window of action of the model was set, if the `data.by.week` parameter is flagged as `TRUE` it returns on the second element of the output list, the summarized data by week. 

```{r no_age_data}
nowcasting_bh_no_age <- nowcasting_inla(dataset = sragBH, 
                                        date_onset = DT_SIN_PRI, 
                                        date_report = DT_DIGITA, 
                                        data.by.week = T)
head(nowcasting_bh_no_age$data)
```

This element is the counts of cases by each delay days. It is known as the delay triangle, if we table the delay amount against the date of onset of first symptoms, we can see the pattern of the delay for the cases.  

```{r delay-triangle}
library(dplyr)

data_triangle <- nowcasting_bh_no_age$data |> 
  filter(delay < 30) |> 
  arrange(delay) |> 
  select(-Time)

data_triangle |> 
  filter(dt_event >= (max(dt_event) - 84),
         delay <= 10) |> 
  tidyr::spread(key = delay, value = Y)

```

We just look at the amount of cases with than 10 weeks of delay or less and 84 days before the latest date. The default maximum is 30 weeks delay considered at nowcasting estimation.  

If this element is groped and summarized by the onset of symptoms date, here `DT_SIN_PRI`, it is the epidemiological curve observed. To example it, we plot the estimate and the epidemiological curve all together.

```{r no_age_plot}
library(ggplot2)

data_by_week <- nowcasting_bh_no_age$data |> 
  dplyr::group_by(dt_event) |> 
  dplyr::reframe(
    observed = sum(Y, na.rm = T)
  ) |>
  dplyr::filter(dt_event >= max(dt_event)-270)

nowcasting_bh_no_age$total |> 
  filter(dt_event >= (max(dt_event)-270)) |> 
  ggplot(aes(x = dt_event, y = Median, col = 'Nowcasting')) +
  geom_line(data = data_by_week, 
            aes(x = dt_event, y = observed, col = 'Observed'))+
  geom_ribbon(aes(ymin = LI, ymax = LS, col = NA), alpha = 0.2, show.legend = F)+
  geom_line()+
  theme_bw()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  scale_color_manual(values = c('grey50', 'black'), name = '')+
  scale_x_date(date_breaks = '2 weeks', date_labels = '%V/%y', name = 'Date in Weeks')+
  labs(x = '', y = 'Nº Cases')
```


## Structured data, Age

The first improvement we have done on the baseline model for nowcasting with a non-structured data is to have the same model for categorical class of the case counts, due to the epidemiological course of SARS-CoV-2, we expect to different ages having different delays distribution. We call this kind of looking into to the data as the structured data, as the data now have identifiers for time, delay and the age class of the cases. To the structured data we fit again a Negative binomial distribution to the cases count at time $t$ with delay $d$. Differently, from the non-structured case the model now gives random effects to the delay distribution and time distribution by each of the age-class chosen by the user to break the data. The model has the form now:

$$\begin{equation}Y_{t,d,a} \sim  NegBinom(\lambda_{t,d,a}, \phi), \\
\log(\lambda_{t,d,a}) =  \alpha_a + \beta_{t,a} + \gamma_{d,a}, \\ 
\beta_{t,a} := u_t - u_{(t-1)} - u_{(t-2)} \sim N(0,\tau_{a, \beta}); \gamma_{d,a} := u_d - u_{(d-1)} \sim N(0, \tau_{a, \gamma}) \\
t=1,2,\ldots,T, \\ 
d=1,2,\ldots,D, \\ 
a=1,2,\ldots,A, \\
\end{equation}$$

where each age class, $a$, has an intercept $\alpha_a$ following a Gaussian distribution with a very large variance, the time-age random effects, $\beta_{t,a}$, follow a joint multivariate Gaussian distribution with a separable variance components an independent Gaussian term for the age classes with precision $\tau_{a,\beta}$ and a second order random walk term for the time with precision $\tau_{\beta}$. Analogously, the delay-age random effects, $\gamma_{d,a}$, follow a joint multivariate Gaussian distribution with a separable variance components an independent Gaussian term for the age classes with precision $\tau_{a,\gamma}$ and a first order random walk term for the time with precision $\tau_{\gamma}$. The model is then completed by INLA default prior distributions for $\phi$, $\tau_{a,\beta}$, $\tau_{a,\gamma}$, $\tau_{a,\beta}$ and $\tau_\gamma$. See [`nbinomial`](https://inla.r-inla-download.org/r-inla.org/doc/likelihood/likelihood-example.pdf), [`rw1`](https://inla.r-inla-download.org/r-inla.org/doc/latent/rw1.pdf) and [`rw2`](https://inla.r-inla-download.org/r-inla.org/doc/latent/rw1.pdf) INLA help pages.

This new model corrects the delay taking into account the effects of age classes and the interactions of each age class between time and also delay. Now the model needs a flag indicating which is the column on the data set which will be used to break the data into age classes and another parameter flagging on how the age classes will be split. This is given by the parameters `age_col` and `bins_age`. We also pass two additional parameters, `data.by.week` to return the epidemiological curve out of window of action of nowcasting estimate and `return.age` to inform we desire a nowcasting result in two ways, the total aggregation estimate and the age-stratified estimate. The calling of the function has the following form:

```{r nowcasting}
nowcasting_bh_age <- nowcasting_inla(dataset = sragBH, 
                                     bins_age = "10 years",
                                     data.by.week = T, 
                                     date_onset = DT_SIN_PRI, 
                                     date_report = DT_DIGITA,
                                     age_col = Idade)
```

Each of the estimates returned by `nowcasting_inla` has the same form as in the non-structured case. On the nowcasting estimates, it returns a `data.frame` with the posterior median and 50% and 95% credible intervals, (LIb and LSb) and (LI and LS) respectively.

```{r plot}
library(ggplot2)

dados_by_week <- nowcasting_bh_age$data |> 
  dplyr::group_by(dt_event) |> 
  dplyr::reframe(
    observed = sum(Y, na.rm = T)
  ) |>
  dplyr::filter(dt_event >= max(dt_event)-270)


nowcasting_bh_age$total |> 
  ggplot()+
  geom_line(aes(x = dt_event, y = Median, 
                col = 'Nowcasting'))+
  geom_line(data = dados_by_week, 
            aes(x = dt_event, y = observed, 
                col = "Observed"))+
  geom_ribbon(aes(x = dt_event, y = Median,
                  ymin = LI, ymax = LS), 
              alpha = 0.2, show.legend = F)+
  theme_bw()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  scale_color_manual(values = c('grey50', 'black'), name = '')+
  scale_x_date(date_breaks = '2 weeks', date_labels = '%V/%y', name = 'Date in Weeks')+
  labs(x = '', y = 'Nº Cases')
```

For sake of completeness we plot both estimates together to check how different they are. 

```{r comparison}


ggplot()+
  geom_line(data = dados_by_week, 
            aes(x = dt_event, y = observed, 
                color = "Observed"))+
  geom_line(data = nowcasting_bh_no_age$total |> 
              filter(dt_event >= (max(dt_event)-270)),
            aes(x = dt_event, y = Median, 
                color = 'Nowcasting - Non-structured'))+
  geom_ribbon(data = nowcasting_bh_no_age$total |> 
              filter(dt_event >= (max(dt_event)-270)),
              aes(x = dt_event, y = Median,
                  ymin = LI, ymax = LS,
                  fill = "Nowcasting - Non-structured"), 
              alpha = 0.5, show.legend = F)+
  geom_line(data = nowcasting_bh_age$total |> 
              filter(dt_event >= (max(dt_event)-270)),
            aes(x = dt_event, y = Median, 
                color = 'Nowcasting - Structured'))+
  geom_ribbon(data = nowcasting_bh_age$total |> 
              filter(dt_event >= (max(dt_event)-270)),
              aes(x = dt_event, y = Median,
                  ymin = LI, ymax = LS,
                  fill = "Nowcasting - Structured"), 
              alpha = 0.5, show.legend = F)+
  theme_bw()+
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90))+
  scale_color_manual(values = c("lightblue4", 
                                "orange4", 
                                "black"),
                     name = "")+
  scale_fill_manual(values = c("lightblue1", 
                                "orange1", 
                                "black"),
                    name = "")+
  scale_x_date(date_breaks = '2 weeks', 
               date_labels = '%V/%y', 
               name = 'Date in Weeks')+
  labs(x = '', 
       y = 'Nº Cases')
```

The nowcasting when using the age information is narrower and shows a less decreasing tendency at the end of the time series. This is due to as each age classes have a nowcasting model acting on it, it can capture effects once mascaraed when not breaking it by age.

## Conclusion

Over this vignette we learned how to use the `nowcasting_inla()` and how `nowcaster` can employ two different models, one without considering differences of delay by age class and another considering differences of delay per age classes. Finally, we compared both estimates, showing that the model that uses the age class information to nowcasting produce narrower estimates
