---
title: "Structured data"
output: rmarkdown::html_vignette
author: "Rafael Lopes & Leonardo Bastos"
vignette: >
  %\VignetteIndexEntry{Structured data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = F,
  message = F,
  echo = T
)
```

As in the Get Started we start by loading the package and its lazy data, by:

```{r data-bh}
library(nowcaster)
# Loading Belo Horizonte SARI dataset
data(sragBH)
```

## Non-structured data

Th Get Started example it is a non-strucutured data estimation, here we give a more detailed on the description of this type of data and how it can change the nowcasting estimation. 

Now we call the nowcasting function, it has by default the parametrization to take the data and estimate with a non-structured data form. The estimate fits a negative binomial distribution, $NegBinom(\lambda_{t,d}, \phi)$, to the cases count at time $t$ with delay $d$, $\phi$ is the dispersion parameter. The rate $\lambda_{t,d}$ is then parameterized in a log-linear format by a constant term added by structured delay random effects and structured time random effects. Hence, the model is given by the following:

$$\begin{equation}
Y_{t,d} \sim NegBinom(\lambda_{t,d}, \phi), \\
\log(\lambda_{t,d}) = \alpha + \beta_t + \gamma_d, \\
t=1,2,\ldots,T, \\ d=1,2,\ldots,D,
\end{equation}$$

where the intercept $\alpha$ follows is Gaussian distribution with a very large variance, $\beta_t$ is follows a second order random walk with precision $\tau_\beta$, $\gamma_d$ a first-order random walk with precision $\tau_\gamma$. The model is then completed by INLA default prior distributions for $\phi$, $\tau_\beta$, and $\tau_\gamma$. See nbinom, rw1 and rw2 INLA help pages.

The call of the function is straightforward, it simply needs a dataset as input, here the `LazyData` loaded in the namespace of the package. The function has 3 mandatory parameters, `dataset` for the parsing of the dataset to be nowcasted, `date_onset` for parsing the column name which is the date of onset of symptoms and `date_report` which parses the column name for the date of report of the cases. Here this columns are "DT_SIN_PRI" and "DT_DIGITA", respectively. 

```{r no_age}
nowcasting_bh_no_age <- nowcasting_inla(dataset = sragBH, 
                                        date_onset = DT_SIN_PRI, 
                                        date_report = DT_DIGITA)
head(nowcasting_bh_no_age$total)
```

This calling will return only the nowcasting estimate and its Confidence Interval (CI) for two different Credible interval, `LIb` and `LSb` are the max and min CI, respectively, with credibility of 50% and `LI` and `LS` are the max and min CI, respectively, with credibility of 95%.

`nowcasting_inla` has the option to return the curve for when the nowcasting estimate was set the window of action of the model, if the `data.by.week` parameter is flagged as `TRUE` it returns on the second element of the output list the summarized data by week. 

```{r no_age_data}
nowcasting_bh_no_age <- nowcasting_inla(dataset = sragBH, 
                                        date_onset = DT_SIN_PRI, 
                                        date_report = DT_DIGITA, 
                                        data.by.week = T)
head(nowcasting_bh_no_age$data)
```
This element it is the counts of cases by the amount of delay. It is known as the delay triangle, if we table the delay amount against the data of onset of first symptoms, it can see how is the pattern of the delay for the cases.  

```{r delay-triangle}
library(dplyr)

data_triangle <- nowcasting_bh_no_age$data |> 
  filter(Delay < 30) |> 
  arrange(Delay)

delay_triangle<-table(data_triangle$date_onset, 
                      data_triangle$Delay,
                      dnn = list("Date of Onset", "Delay"))
delay_triangle
```

We just look at the amount of cases with 30 weeks of delay or less, it is the default maximum delay considered at nowcasting estimation.  

If this element is groped by and summarized by the onset of symptoms date, here `DT_SIN_PRI`, it is the epidemiological curve observed. To example it, we plot the estimate and the epidemiological curve all together.

```{r no_age_plot}
library(ggplot2)

data_by_week <- nowcasting_bh_no_age$data |>
  ungroup() |> 
  filter(date_onset >= (max(date_onset)-270))

nowcasting_bh_no_age$total |> 
  filter(dt_event >= (max(dt_event)-270)) |> 
  ggplot(aes(x = dt_event, y = Median, col = 'Nowcasting')) +
  geom_line(data = data_by_week, aes(date_onset, y = observed, col = 'Observed'))+
  geom_ribbon(aes(ymin = LI, ymax = LS, col = NA), alpha = 0.2, show.legend = F)+
  geom_line()+
  theme_bw()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  scale_color_manual(values = c('grey50', 'black'), name = '')+
  scale_x_date(date_breaks = '2 weeks', date_labels = '%V/%y', name = 'Date in Weeks')+
  labs(x = '', y = 'Nº Cases')
```


## Structured data, Age

For the structured data the `nowcasting_inla()` fits again a Negative binomial distribution to the cases count at time $t$ with delay $d$. Differently, from the non-structured case the model now gives random effects to the delay distribution and and time distribution by each of the age-class chosen by the user to break the data. The model has the form now:

$$\begin{equation}Y_{t,d,a} \sim  NegBinom(\lambda_{t,d,a}, \phi), \\
\log(\lambda_{t,d,a}) =  \alpha_a + \beta_{t,a} + \gamma_{d,a}, \\ \quad t=1,2,\ldots,T, \\ d=1,2,\ldots,D, \\ a=1,2,\ldots,A, \end{equation}$$

where each age class, $a$, has an intercept $\alpha_a$ following a Gaussian distribution with a very large variance, the time-age random effects, $\beta_{t,a}$, follow a joint multivariate Gaussian distribution with a separable variance components an independent Gaussian term for the age classes with precision $\tau_{age,\beta}$ and a second order random walk term for the time with precision $\tau_{\beta}$. Analogously, the delay-age random effects, $\gamma_{d,a}$, follow a joint multivariate Gaussian distribution with a separable variance components an independent Gaussian term for the age classes with precision $\tau_{age,\gamma}$ and a first order random walk term for the time with precision $\tau_{\gamma}$. The model is then completed by INLA default prior distributions for $\phi$, $\tau_{age,\beta}$, $\tau_{age,\gamma}$, $\tau_{\beta}$ and $\tau_\gamma$. See nbinom, iid, rw1 and rw2 INLA help pages.

This new model corrects the delay taking into account the effects of age classes and the interactions of each age class between time and also delay. Now the model needs a flag indicating which is the column on the dataset which will be used to break the data into age classes and how the age classes will be split. This is given by the parameters `age_col` and `bins_age`. We pass three additional parameters, `data.by.week` to return the epidemiological curve out of window of action of nowcasting estimate and `return.age` to inform we desire a nowcasting result in two ways, the total aggregation estimate and the age-stratified estimate. The calling of the function has the following form:

```{r nowcasting}
nowcasting_bh_age <- nowcasting_inla(dataset = sragBH, 
                                   bins_age = "10 years",
                                   data.by.week = T, 
                                   date_onset = DT_SIN_PRI, 
                                   date_report = DT_DIGITA,
                                   age_col = Idade)
```

Each of the estimates returned by `nowcasting_inla` has the same form as in the non-structured case. On the nowcasting estimates, it returns a data.frame with the posterior edian and 50% and 95% credible intervals, (LIb and LSb) and (LI and LS) respectively.

```{r plot}
library(ggplot2)

dados_by_week <- nowcasting_bh_age$data |> 
  ungroup() |> 
  filter(date_onset >= (max(date_onset)-270))

nowcasting_bh_age$total |> 
  ggplot(aes(x = dt_event, y = Median, col = 'Median'))+
  geom_line()+
  geom_line(data = dados_by_week, aes(date_onset, y = observed))+
  geom_ribbon(aes(ymin = LI, ymax = LS, col = 'IC'), alpha = 0.2)+
  theme_bw()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  scale_color_manual(values = c('grey90', 'black'), name = '')+
  scale_x_date(date_breaks = '2 weeks', date_labels = '%V/%y', name = 'Date in Weeks')+
  labs(x = '', y = 'Nº Cases')
```

